### 2023-05-13 misc/expo/additive.tex
Roth's thm lemma 1

If $\phi : A\to B$ is a $k$-homorphism with an inverse that is also a $k$-homorphism, then we say that
$\phi$ is a {\it Freiman isomorphism of order $k$} or, more briefly, a {\it $k$-isomorphism}. These
maps preserve a lot of the properties that are of interest in additive combinatorics. Suppose that $A$
is $k$-isomorphic to $B$. Then
\medskip
\item{i)} $|kA| = |kB|$;
\smallskip
\item{ii)} $|rA-sA| = |rB-sB|$ for all $r$ and $s$ with $r+s\le k$;
\smallskip
\item{iii)} $E(A) = E(B)$ if $k=2$; and
\smallskip
\item{iv)} if $k=2$ and $A$ contains an arithmetic progression of length $r$, then so does $B$.
\medskip
Of these statements, the only slightly nontrivial one is (iv). It follows from the fact that an
arithmetic progression is a sequence $(a_1,\ldots,a_r)$ with $a_i+a_i = a_{i-1}+a_{i+1}$ for all $1<i<r$.
Taking $\phi$ of both sides gives
$$\phi(a_i) + \phi(a_i) = \phi(a_i+a_i) = \phi(a_{i-1}+a_{i+1}) = \phi(a_{i-1}) + \phi(a_{i+1}).$$
Let us now state and prove two important ``embedding lemmas''. These lemmas tell us that if
$kA-kA$ is not too large, then the additive set $A$ is $k$-isomorphic to a subset of a group that
is simpler, in some sense.

@@

Let $G$ be a finite abelian group, let $\chi_1, \ldots, \chi_k$ be characters on $G$, and let
$\delta>0$. The {\it Bohr set} $B(\chi_1,\ldots,\chi_k;\delta)$ is the set
$$\bigl\{ x\in G : \chi_i(x)\in e\bigl([-\delta,\delta]\bigr)\ \hbox{for all}\ 1\le i\le k\bigr\}.$$
(As before, we write $e(\alpha)$ for $e^{2\pi i\alpha}$.) We say that such a Bohr set has {\it dimension} $k$
and {\it width} $\delta$.

If $G= \FF_p^n$, we can write $\chi_i = \chi_{u_i}$ where $\chi_{u_i}(x) = e(u_i\cdot x/p$. If $\delta<1/p$,
then $\chi_{u_i}(x)\in e\bigl([-\delta,\delta]\bigr)$ if and only if
$(u_i\cdot x) / p \bmod 1\in [-\delta,\delta]$. But $u_i\cdot x$ is an integer, so this is true if and only
if $u_i\cdot x = 0$, so consequently
$$B(\chi_1,\ldots,\chi_k;\delta) = \{ x\in \FF_p^n : u_i\cdot x = 0\ \hbox{for all}\ 1\le i\le k\},$$
which is a subspace cut out by $k$ linear conditions, i.e., of codimension at most $k$.

\parenproclaim Lemma {\advthm} (Bogolyubov). Let $G$ be a finite abelian group. Let $A\subseteq G$ be a
subset of density $\alpha$. Then $2A-2A$ contains a Bohr set of dimension at most $1/\alpha^2$ and width
$1/4$.

\proof For $x\in G$, let
$$\eqalign{
f(x) &= \bigl( \one_A * \one_A * (\mone_A) * (\mone_A)\bigr)(x) \cr
&= \ex_{p+q-r-s=x} \one_A(p)\one_A(q)\one_A(r)\one_A(s).
}$$
Then $x\in 2A-2A$ if and only if $f(x)\ne 0$. We want to find a Bohr set on which $f$ does not vanish.

Note first that for any character $\chi$,
$$\hat{\mone_A}(\chi) = \ex_x \one_A(-x)\bar{\chi(x)} = \ex_x \one_A(x)\bar{\chi(-x)} = \ex_x\one_A(x)\chi(x)
=\bar{\hat{\one_A}(\chi)}.$$
(The last identity relies on the fact that $\one_A(x)$ is real.) This implies that
$\hat{\one_A}(\hat{\mone_A}) = |\one_A|^2$. Now we apply the Fourier inversion
formula and the convolution law to expand
$$\eqalign{
f(x) &= \sum_\chi \hat{\one_A*\one_A*(\mone_A)*(\mone_A)}(\chi)(\chi)(x) \cr
&= \sum_\chi \hat{\one_A}(\chi)^2 \hat{\mone_A}(\chi)^2 \chi(x) \cr
&= \sum_\chi \bigl| \hat{\one_A}(\chi) \bigr|^4 \chi(x). \cr
}$$
Let
$$K = \bigl\{ \chi\in \hat G : \bigl| \hat{\one_A}(\chi) \bigr| \ge \alpha^{3/2}\bigr\}$$
and let $B$ be the Bohr set $B(K;1/4)$. Then the expression above for $f(x)$ can be split into
$$f(x) =
\bigl| \hat{\one_A}(\chi_0) \bigr|^4
+ \sum_{\chi\in K} \bigl| \hat{\one_A}(\chi) \bigr|^4 \chi(x)
+ \sum_{\chi\notin K\cup \{\chi_0\}} \bigl| \hat{\one_A}(\chi) \bigr|^4 \chi(x).$$
We shall assume that $x\in B$ and deal with each of the three terms separately.

The first term is easy; since $A$ has density $\alpha$, $\hat{\one_A}(\chi_0) = \alpha$, and thus the first
time is $\alpha^4$. For the second term, note that if $\chi\in K$ and $x\in B$, then
$\chi(x)\in e\bigl([-1/4, 1/4]\bigr)$. This means that $\chi(x)$ has argument between $-\pi/2$ and $\pi/2$,
and {\it a fortiori} has nonnegative real part. Now, if $\chi\notin K$ and $\chi\ne \chi_0$,
then $\bigl|\hat{\one_A}\bigr| < \alpha^{3/2}$, so we take the absolute value of the third term
and apply the triangle inequality and Parseval's identity to obtain
$$\Bigl|\sum_{\chi\notin K\cup \{\chi_0\}} \bigl| \hat{\one_A}(\chi) \bigr|^4 \chi(x)\Bigr|
\le \sum_{\chi\notin K\cup\{\chi_0\}} \bigl|\hat{\one_A}(\chi)\bigr|^4
< \sum_{\chi\in \hat G} \alpha^3 \bigl| \hat{\one_A}(\chi)\bigr|^2
= \alpha^3\langle \hat{\one_A},\hat{\one_A}\rangle.$$
But by Parseval's identity, the right-hand side equals $\alpha^3 \langle \one_A,\one_A\rangle = \alpha^4$,
so we find that the real part of the third term is more than $-\alpha^4$. This means that $\Re f(x) > 0$,
so $f(x)\ne 0$ for all $x\in B$; that is, $B\subseteq 2A-2A$.

It remains to give an upper bound on the size of $K$. We have
$$\alpha = \sum_{\chi\in \hat G} \bigl| \one_A(\chi)\bigr|^2 \ge
\sum_{\chi\in K} \bigl|\hat{\one_A(\chi)}\bigr|^2 \ge |K|\alpha^3,$$
meaning that $|K|\le 1/\alpha^2$.\slug

When $A$ is a subset of density $\alpha$ in $G = \FF_p^n$, Bogolyubov's lemma tells us that $2A-2A$ contains
a subspace of codimension at most $1/\alpha^2$.

>>>>

If $\phi : A\to B$ is a $k$-homomorphism with an inverse that is also a $k$-homomorphism, then we say that
$\phi$ is a {\it Freiman isomorphism of order $k$} or, more briefly, a {\it $k$-isomorphism}. These
maps preserve a lot of the properties that are of interest in additive combinatorics. Suppose that $A$
is $k$-isomorphic to $B$. Then
\medskip
\item{i)} $|kA| = |kB|$;
\smallskip
\item{ii)} $|rA-sA| = |rB-sB|$ for all $r$ and $s$ with $r+s\le k$;
\smallskip
\item{iii)} $E(A) = E(B)$ if $k=2$; and
\smallskip
\item{iv)} if $k=2$ and $A$ contains an arithmetic progression of length $r$, then so does $B$.
\medskip
Of these statements, the only slightly nontrivial one is (iv). It follows from the fact that an
arithmetic progression is a sequence $(a_1,\ldots,a_r)$ with $a_i+a_i = a_{i-1}+a_{i+1}$ for all $1<i<r$.
Taking $\phi$ of both sides gives
$$\phi(a_i) + \phi(a_i) = \phi(a_i+a_i) = \phi(a_{i-1}+a_{i+1}) = \phi(a_{i-1}) + \phi(a_{i+1}).$$
Let us now state and prove two important ``embedding lemmas''. These lemmas tell us that if
$kA-kA$ is not too large, then the additive set $A$ is $k$-isomorphic to a subset of a group that
is simpler, in some sense.

@@

Let $Z$ be a finite abelian group, let $\chi_1, \ldots, \chi_k$ be characters on $Z$, and let
$\delta>0$. The {\it Bohr set} $B(\chi_1,\ldots,\chi_k;\delta)$ is the set
$$\bigl\{ x\in Z : \chi_i(x)\in e\bigl([-\delta,\delta]\bigr)\ \hbox{for all}\ 1\le i\le k\bigr\}.$$
(As before, we write $e(\alpha)$ for $e^{2\pi i\alpha}$.) We say that such a Bohr set has {\it dimension} $k$
and {\it width} $\delta$.

If $Z= \FF_p^n$, we can write $\chi_i = \chi_{u_i}$ where $\chi_{u_i}(x) = e(u_i\cdot x/p)$. If $\delta<1/p$,
then $\chi_{u_i}(x)\in e\bigl([-\delta,\delta]\bigr)$ if and only if
$(u_i\cdot x) / p \bmod 1\in [-\delta,\delta]$. But $u_i\cdot x$ is an integer, so this is true if and only
if $u_i\cdot x = 0$, so consequently
$$B(\chi_1,\ldots,\chi_k;\delta) = \{ x\in \FF_p^n : u_i\cdot x = 0\ \hbox{for all}\ 1\le i\le k\},$$
which is a subspace cut out by $k$ linear conditions, i.e., of codimension at most $k$.

\parenproclaim Lemma {\advthm} (Bogolyubov). Let $Z$ be a finite abelian group. Let $A\subseteq Z$ be a
subset of density $\alpha$. Then $2A-2A$ contains a Bohr set of dimension at most $1/\alpha^2$ and width
$1/4$.

\proof For $x\in Z$, let
$$\eqalign{
f(x) &= \bigl( \one_A * \one_A * (\mone_A) * (\mone_A)\bigr)(x) \cr
&= \ex_{p+q-r-s=x} \one_A(p)\one_A(q)\one_A(r)\one_A(s).
}$$
Then $x\in 2A-2A$ if and only if $f(x)\ne 0$. We want to find a Bohr set on which $f$ does not vanish.

Note first that for any character $\chi$,
$$\hat{\mone_A}(\chi) = \ex_x \one_A(-x)\bar{\chi(x)} = \ex_x \one_A(x)\bar{\chi(-x)} = \ex_x\one_A(x)\chi(x)
=\bar{\hat{\one_A}(\chi)}.$$
(The last identity relies on the fact that $\one_A(x)$ is real.) This implies that
$\hat{\one_A}(\hat{\mone_A}) = |\one_A|^2$. Now we apply the Fourier inversion
formula and the convolution law to expand
$$\eqalign{
f(x) &= \sum_\chi \hat{\one_A*\one_A*(\mone_A)*(\mone_A)}(\chi)(\chi)(x) \cr
&= \sum_\chi \hat{\one_A}(\chi)^2 \hat{\mone_A}(\chi)^2 \chi(x) \cr
&= \sum_\chi \bigl| \hat{\one_A}(\chi) \bigr|^4 \chi(x). \cr
}$$
Let
$$K = \bigl\{ \chi\in \hat Z : \bigl| \hat{\one_A}(\chi) \bigr| \ge \alpha^{3/2}\bigr\}$$
and let $B$ be the Bohr set $B(K;1/4)$. Then the expression above for $f(x)$ can be split into
$$f(x) =
\bigl| \hat{\one_A}(\chi_0) \bigr|^4
+ \sum_{\chi\in K} \bigl| \hat{\one_A}(\chi) \bigr|^4 \chi(x)
+ \sum_{\chi\notin K\cup \{\chi_0\}} \bigl| \hat{\one_A}(\chi) \bigr|^4 \chi(x).$$
We shall assume that $x\in B$ and deal with each of the three terms separately.

The first term is easy; since $A$ has density $\alpha$, $\hat{\one_A}(\chi_0) = \alpha$, and thus the first
time is $\alpha^4$. For the second term, note that if $\chi\in K$ and $x\in B$, then
$\chi(x)\in e\bigl([-1/4, 1/4]\bigr)$. This means that $\chi(x)$ has argument between $-\pi/2$ and $\pi/2$,
and {\it a fortiori} has nonnegative real part. Now, if $\chi\notin K$ and $\chi\ne \chi_0$,
then $\bigl|\hat{\one_A}\bigr| < \alpha^{3/2}$, so we take the absolute value of the third term
and apply the triangle inequality and Parseval's identity to obtain
$$\Bigl|\sum_{\chi\notin K\cup \{\chi_0\}} \bigl| \hat{\one_A}(\chi) \bigr|^4 \chi(x)\Bigr|
\le \sum_{\chi\notin K\cup\{\chi_0\}} \bigl|\hat{\one_A}(\chi)\bigr|^4
< \sum_{\chi\in \hat Z} \alpha^3 \bigl| \hat{\one_A}(\chi)\bigr|^2
= \alpha^3\langle \hat{\one_A},\hat{\one_A}\rangle.$$
But by Parseval's identity, the right-hand side equals $\alpha^3 \langle \one_A,\one_A\rangle = \alpha^4$,
so we find that the real part of the third term is more than $-\alpha^4$. This means that $\Re f(x) > 0$,
so $f(x)\ne 0$ for all $x\in B$; that is, $B\subseteq 2A-2A$.

It remains to give an upper bound on the size of $K$. We have
$$\alpha = \sum_{\chi\in \hat Z} \bigl| \one_A(\chi)\bigr|^2 \ge
\sum_{\chi\in K} \bigl|\hat{\one_A(\chi)}\bigr|^2 \ge |K|\alpha^3,$$
meaning that $|K|\le 1/\alpha^2$.\slug

When $A$ is a subset of density $\alpha$ in $Z = \FF_p^n$, Bogolyubov's lemma tells us that $2A-2A$ contains
a subspace of codimension at most $1/\alpha^2$.

\advsect Freiman's theorem in vector spaces over fields of prime order

Let $Z$ be an additive group and let $A$ be a finite subset of $Z$. It is possible that $|A+A| = |A|$,
but this happens if and only if $A$ is a coset of a subgroup of $Z$ (which means that $Z$ must be finite).
If $A$ is a finite subset of integers, then without loss of generality we can shift $A$ so that $\min A = 0$.
if $A$ only contains $0$ then $|A+A|$ is $1 = 2|A|-1$, and if $|A| > 1$, then let $m = \max A$ and
note that $A+A$ contains $A$ and $m+A$, and $A\cap (m+A) = \{m\}$, so in this case we also have $|A+A|\ge 2|A|-1$.

It is not hard to see that if $A$ is an arithmetic progression
in the integers, then $A+A$ is exactly $2|A|-1$,
but for instance $A = \{1,\ldots,n\}\subseteq \ZZ$ is not contained in any nontrivial subgroup of $\ZZ$ and
it also has $|A+A| = 2|A|-1$.
So next we might amend our conjecture to say that if $|A+A|$ is small, then $A$ is a large subset of an arithmetic
progression. But this is also false. Consider the set
$$\{0,1,2,\quad 10,11,12,\quad 20,21,22,\quad 30,31,32\}.$$
In some sense this is a ``progression of progressions,'' or a $2$-dimensional arithmetic progression. Such
a set is the projection of some rectangle in $\ZZ^2$, and from this observation it is not hard to see
that $|A+A|\le 4|A|$. If we increase $d$, we can still say that $|A+A|\le 2^d|A|$.

Once we add these $d$-dimensional arithmetic progressions into consideration, it turns out that we can have
an inverse statement. Concretely, if $|A+A|\le C|A|$, then there exists a $d$-dimensional arithmetic
progression $P$ such that $|P|\le K|A|$ and $A\subseteq P$, where $d$ and $K$ depend only on $C$ (otherwise
we could simply embed $A$ into $[\min A, \max A]$).

This statement is called Freiman's theorem, as it was first proved by Freiman over the integers.
In these notes we will prove an analogue of it for subsets of $\FF_p^N$, in which ``$d$-dimensional arithmetic
progression'' is replaced by ``subspace''. First we need a lemma.

\parenproclaim Lemma {\advthm} (Ruzsa covering lemma). Let $Z$ be an abelian group and let $A$ and
$B$ be finite subsets of $Z$. Then there is a set $K\subseteq A$ of size at most $|A+B|/|B|$ such that
$A\subseteq K+B-B$.

\proof Let $K = \{a_1,\ldots,a_k\}$ be a maximal subset of $A$ with the property that the sets $a_i+B$
are disjoint. For all $a\in A$ there is $i$ such that $(a+B) \cap (a_i+B) \ne \emptyset$, otherwise we
could add $a$ to $K$ and contradict maximality. Equivalently, we can find $b$ and $b'$ such that
$a+b = a_i + b'$ so $a\in a_i + B - B$. Hence $A\subseteq K+B-B$. Lastly, $|K|\cdot |B| = |K+B| \le |A+B|$.\slug

Next, we prove two simple lemmas about Freiman isomorphisms, the second concerning
how they behave on subspaces of $\FF_p^n$.

\edef\leminducedfreiman{\the\thmcount}
\proclaim Lemma {\advthm}. A Freiman isomorphism of order $k(r+s)$ from $A$ to $B$ induces a Freiman
isomorphism of order $k$ from $rA-sA$ to $rB-sB$.

\proof This proof is conceptually very easy but notationally a total nightmare.
Let $\phi$ be a $k(r+s)$ isomorphism from $A$ to $B$. Then
$$x_1+\cdots x_{k(r+s)} = y_1+\cdots+y_{k(r+s)}$$
if and only if
$$\phi(x_1)+\cdots \phi(x_{k(r+s)}) = \phi(y_1)+\cdots+\phi(y_{k(r+s)})$$
Now let $g_1,\ldots,g_k, h_1,\ldots,h_k$ be $2k$ elements in $rA-sA$. Then the expression
$$g_1+\cdots g_k = h_1+\cdots+h_k$$
has a sum of $kr$ elements of $A$ and $ks$ elements of $-A$ on each side. We can subtract
off the elements of $-A$ on each side to get $k(r+s)$ elements of $A$ on either side, and the equality
holds if and only if it holds with $\phi$ applied to every term. So letting
$$g_i = g'_{i,1}+\cdots+g'_{i,r}-g''_{i,1}-\cdots-g''_{i,s}$$
and
$$h_i = h'_{i,1}+\cdots+h'_{i,r}-h''_{i,1}-\cdots-h''_{i,s}$$
for all $i$, we find that $g_1+\cdots g_k = h_1+\cdots+h_k$ holds if and only if
$$\eqalign{
\phi(g'_{1,1})+\cdots+&\phi(g'_{k,r}) + \phi(h''_{1,1})+\cdots+\phi(h''_{k,s}) \cr
&= \phi(h'_{1,1})+\cdots+\phi(h'_{k,r}) + \phi(g''_{1,1})+\cdots+\phi(g''_{k,s}).
}$$
Subtracting all the double-prime terms from both sides yields
$$\eqalign{
\phi(g'_{1,1})+\cdots+&\phi(g'_{k,r}) - \phi(g''_{1,1})-\cdots-\phi(g''_{k,s}) \cr
&= \phi(h'_{1,1})+\cdots+\phi(h'_{k,r})  - \phi(h''_{1,1})-\cdots-\phi(h''_{k,s}),
}$$
and letting
$$\psi(g_i) = \phi(g'_{i,1}) + \cdots + \phi(g'_{i,r}) - \phi(g''_{i,1}) -\cdots-\phi(g''_{i,s})$$
and
$$\psi(h_i) = \phi(h'_{i,1}) + \cdots + \phi(h'_{i,r}) - \phi(h''_{i,1}) -\cdots-\phi(h''_{i,s})$$
for all $i$, we see that $g_1+\cdots g_k = h_1+\cdots+h_k$ holds if and only if
$$\psi(g_1)+\cdots \psi(g_k) = \psi(h_1)+\cdots+\psi(h_k).$$
It is invertible, since we can just apply $\phi^{-1}$ termwise to an element of $rB-sB$, so
this is the $k$-isomorphism we need between $rA-sA$ and $rB-sB$.\slug

\edef\lemaffinefreiman{\the\thmcount}
\proclaim Lemma {\advthm}. Let $X$ be a linear subspace of $\FF_p^n$ and let $\phi$ be a $2$-isomorphism
from $X$ to a set $A\subseteq \FF_p^n$. Then $A$ is an affine subspace of $\FF_p^n$ of dimension $\dim X$.

\proof We shall show that $Y = A - \phi(0)$ is a linear subspace of $\FF_p^n$. Clearly it contains $0$,
since $A$ contains $\phi(0)$. Now let $y_1,y_2\in Y$ so that $y_1 + \phi(0)$ and $y_2+\phi(0)$ are both in $A$.
Let $x_1 = \phi^{-1}\bigl(y_1+\phi(0)\bigr)$, $x_2 = \phi^{-1}\bigl(y_1+\phi(0)\bigr)$, and let $x$ denote
their sum in $X$. Then since $x_1 + x_2 = x+0$, we have
$$y_1 + \phi(0) + y_2 + \phi(0) = \phi(x) + \phi(0).$$
Subtracting $\phi(0)$ from both sides, we find that $y_1 + y_2 + \phi(0) = \phi(x)$, meaning
that $y_1 + y_2 \in A - \phi(0) = Y$. This proves that $Y$ is closed under finite sums, and since scalar
multiplication over a finite field can always be re\"expressed as a finite sum, we find that
$A$ is an affine subspace of $\FF_p^n$ in bijection with $X$, that is, of dimension $\dim X$.\slug

We are now able to prove Freiman's theorem for $\FF_p^n$, whose proof amounts to little more than
a concatenation of several previous results.

\parenproclaim Theorem {\advthm} (Freiman's theorem in $\FF_p^n$). Let $A\subseteq \FF_p^n$ and suppose
that $|A+A|\le C|A|$. Then there is a subspace $X$ of $\FF_p^n$ such that $A\subseteq X$ and $|X|\le C'|A|$,
where $C'$ depends only on $C$ and $p$.

\proof By Pl\"unnecke's theorem, $|8A-8A| \le C^{16} |A|$, hence by Ruzsa's embedding lemma, $A$
is $8$-isomorphic to a subset $A_1\subseteq \FF_p^m$ where $p^m \le pC^{16}|A|$. By Bogolyubov's lemma
with $\alpha = p^{-1}C^{-16}$, $2A_1 - 2A_1$ contains a Bohr set $B(K;1/4)$ of dimension at most
$p^2C^{32}$. But letting $\delta < 1/(2p) \le 1/4$, and $X = B(K;\delta)$, we have $X\subseteq B(K;1/4)$
and $X$ is a linear subspace of $\FF_p^m$ of codimension at most $p^2C^{32}$.

Now since $A$ is $8$-isomorphic to $A_1$, we can apply Lemma~{\leminducedfreiman} to deduce that
$2A-2A$ is $2$-isomorphic to $2A_1 - 2A_1$. It follows that
$2A-2A$ contains a set $Y$ that is $2$-isomorphic to the linear subspace $X$. By Lemma~{\lemaffinefreiman},
this set $Y$ is an affine subspace, of dimension at least $m-p^2C^{32}$, and thus of cardinality
at least $p^{m-p^2C^{32}} \ge p^{p^2C^{32}}|A|$ (here we used the fact that $A_1\subseteq \FF_p^m$ and
$A_1$ is Freiman isomorphic to $A$).

Ruzsa's covering lemma tells us that there exists $L$ of size at most $|A+Y|/|Y|$ such that $A\subseteq L+Y-Y$.
But since $Y$ is an affine subspace, $Y-Y$ is a linear subspace; setting $V = Y-Y$, we have
$$|V| = |Y| \le |2A-2A| \le C^4|A|,$$
which follows from another application of Pl\"unnecke's theorem.
Then by possibly adding $|L|$ basis vectors, $L+V$ is contained in a linear subspace of size at most $p^{|L|} |V|
\le p^{|L|} C^4|A|$.
On the other hand, $L$ has size at most
$${|A+2A-2A|\over |V|} \le {C^5|A| \over p^{-p^2C^{32}} |A|} = C^5p^{p^2C^{32}},$$
where yet again we have used Pl\"unnecke's theorem.
Hence we conclude that $A$ is contained in a linear subspace of size at most
$$ p^{C^5 p^{p^2C^{32}}} C^4|A|.\noskipslug$$

\advsect Roth's theorem

In the previous section we showed that any set $A$ with $A+A$ small
must be a dense subset of a structured set. We didn't prove this over $\ZZ$, but we saw briefly that
the notion of ``structured set'' in that case was a $d$-dimensional arithmetic progression. Now we switch
gears and try to find arithmetic progressions in subsets of $\ZZ$. The first major theorem in this vein was
proved by B.~L.~van der Waerden in 1927.

\parenproclaim Theorem W (van der Waerden, {\rm 1927}). For all $k$ and $r$, there is $n$ such
that if we partition $[n]$ into disjoint sets $C_1\cup \cdots \cup C_r$ then some $C_i$ contains an
arithmetic progression of length $k$.\slug

A stronger statement is given by the celebrated 1975 theorem of E.~Szemer\'edi.

\parenproclaim Theorem S (Szemer\'edi, {\rm 1975}). For every $\delta>0$ and $k\in \NN$ there
exists $n$ such that if $A\subseteq [n]$ and $|A|\ge \delta n$, then $A$ contains an arithmetic progression
of length $k$.\slug

By setting $\delta = 1/r$ and invoking the pigeonhole principle, we see that
Szemer\'edi's theorem implies van der Waerden's theorem. We will not prove either of these statements in
these notes. The first has a relatively elementary proof, but not one that is very additive-combinatorial
in nature, while the second is far too complex to prove in full generality here. We will, however,
prove Szemer\'edi's theorem in the case $k=3$. This special case is called Roth's theorem as it was
proved by K.~Roth in 1953.
Before we get there we first establish three lemmas. In their proofs we will write, e.g., $A$ instead
of $\one_A$, and identify
$r$ with its corresponding element $\chi_r\in \hat{\ZZ_N}$, writing $\hat f(r)$ instead of $\hat f(\chi_r)$.

\proclaim Lemma {\advthm}. Let $N$ be odd and let $A$, $B$, and $C$ be subsets of $\ZZ_N$ with densities
$\alpha$, $\beta$, and $\gamma$ respectively. Suppose that $N > 2/(\alpha\beta\gamma)$. Then either
there exists $x$ and $d\ne 0$ such that $(x,x+d,x+2d)\in A\times B\times C$, or there exists $r\ne 0$
such that $|\hat A(r)| \ge \alpha \beta^{1/2} \gamma^{1/2} / 2$.

\proof Consider the quantity
$$\eqalign{
\ex_{x,d} A(x)B(x+d)C(x+2d) &= \ex_{x+z=2y} A(x)C(z)B(y) \cr
&= \ex_u \bigl( \ex_{x+z=u} A(x)C(z)\bigr) B(u/2).
}$$
Using the notation $f_a(x) = f(x/a)$ for a dilation, we see that this expectation is actually the
inner product $\langle A *C, B_2\rangle$, which by Parseval's identity is equal to
$\langle \hat{A*C}, \hat{B_2}\rangle$.
Recall that the characters of $\ZZ_N$ are given by $\chi_a(x) = e(ax/N)$. Applying the dilation rule gives
$$ \hat{B_2}(r) = \hat{B_2}(\chi_r) = \hat B({\chi_r}^2) = \hat B(\chi_{2r})=\hat B(2r)=(\hat B)_{2^{-1}}(r)$$
where $2^{-1}$ denotes the inverse of $2$ modulo $N$.
This, combined with the convolution law, gives
$\langle \hat{A*C}, \hat{B_2}\rangle = \bigl\langle \hat A\hat C, (\hat B)_{2^{-1}}\bigr\rangle$.
But since $B$ is real-valued,
$$\hat B(-r) = \ex_x B(x)\bar{\chi_{-r}(x)} = \ex_x B(x)\chi_r(x) = \bar{\ex_x B(x)\bar{\chi_r(x)}}
= \bar{\hat B(r)},$$
so putting everything together yields
$$\eqalign{
\ex_{x,d} A(x)B(x+d)C(x+2d) &= \bigl\langle \hat A\hat C, (\hat B)_{2^{-1}}\bigr\rangle \cr
&= \sum_r \hat A(r)\hat C(r) \bar{\hat B(2r)} \cr
&= \sum_r \hat A(r)\hat C(r) \hat B(-2r) \cr
&= \alpha\beta\gamma + \sum_{r\ne 0} \hat A(r) \hat C(r) \hat B(-2r). \cr
}$$
Note that by the triangle and Cauchy--Schwarz inequalities,
$$\eqalign{
\Bigl|\sum_{r\ne 0} \hat A(r) \hat C(r) \hat B(-2r)\Bigr|
&\le \max_{r\ne 0} \bigl|\hat A(r)\bigr| \sum_{r\in \ZZ_N} \bigl|\hat B(-2r)\bigr|\cdot\bigl|\hat C(r)\bigr| \cr
&\le \max_{r\ne 0} \bigl|\hat A(r)\bigr|
\Bigl(\sum_{r\in \ZZ_N} \bigl|\hat B(-2r)\bigr|^2\Bigr)^{1/2}
\Bigl(\sum_{r\in \ZZ_N} \bigl|\hat C(r)\bigr|^2\Bigr)^{1/2} \cr
&= \bignorm{\hat B}_2 \bignorm{\hat C}_2 \max_{r\ne 0} \bigl| \hat A(r)\bigr| \cr
&= \beta^{1/2} \gamma^{1/2} \max_{r\ne 0} \bigl| \hat A(r)\bigr|.\cr
}$$
(We used the fact that $N$ is odd in the third line above.) So
$$
\ex_{x,d} A(x)B(x+d)C(x+2d) \ge \alpha\beta\gamma
-\beta^{1/2} \gamma^{1/2} \max_{r\ne 0} \bigl| \hat A(r)\bigr|,$$
meaning that if $\max_{r\ge 0} \bigl| \hat A(r)\bigr| < \alpha \beta^{1/2}\gamma^{1/2} / 2$, then
$\ex_{x,d} A(x)B(x+d)C(x+2d) > \alpha\beta\gamma/2$. The further assumption that $\alpha\beta\gamma/2 > 1/N$
allows us to conclude that there is some $x$ for which $\ex_{d} A(x)B(x+d)C(x+2d) > 1$, and thus some
$d\ge 0$ such that $(x,x+d,x+2d)\in A\times B\times C$.\slug
